### 2024-11-25 paper/paper.qmd
let eval: false to avoid errors

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - Denise Chang
thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
#| eval: false

library(tidyverse)
library(palmerpenguins)
```

# Introduction

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

## Overview

We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...

Overview text

## Measurement

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Outcome variables

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false
#| eval: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false
#| eval: false

analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: false
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: false
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: false
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: false
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References

### 2024-11-25 scripts/02-download_data.R
update preamble and save as xlsx files

@@ -1,103 +1,42 @@
 #### Preamble ####
-# Purpose: Downloads and saves the data from Open Data Toronto
+# Purpose: Download and save the Paramedic Services Incident Data from Open Data Toronto
 # Author: Denise Chang
-# Date: 20 November 2024
-# Contact: dede.chang@utoronto.ca
+# Date: 25 November 2024
+# Contact: dede.chang@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: None
 
-#### Preamble ####
-# Purpose: Download and save the actual Paramedic Services Incident Data from 2017 to 2022
-# from Open Data Toronto
-# Author: Julia Lee
-# Date: 23 September 2024
-# Contact: jlee.lee@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: None
-# Any other information needed? N/A
-
-
-#### Setting Up the Workspace ####
-
+#### Workspace setup ####
 library(opendatatoronto)
 library(tidyverse)
-library(dplyr)
+library(openxlsx)
 
-#### Downloading the Data #### 
-
+#### Download data ####
+# get package
 package <- show_package("c21f3bd1-e016-4469-abf5-c58bb8e8b5ce")
 
-resources <- list_package_resources("c21f3bd1-e016-4469-abf5-c58bb8e8b5ce")
+resources <-
+  list_package_resources("c21f3bd1-e016-4469-abf5-c58bb8e8b5ce")
 
-datastore_resources <- filter(resources, tolower(format) %in% c('xlsx', 'geojson'))
+xlsx_resource <-
+  resources[resources$name == "paramedic-services-incident-data-2017-2022",]
 
-data <- filter(datastore_resources, row_number()==1) %>% 
-  get_resource()
+raw_data <- get_resource(xlsx_resource)
 
-data
+#### Save data ####
 
-#### Creating a Dataframe with the Downloaded Data ####
+# Each year is saved as a separate file to preserve the structure.
+data_2017 <- tibble(raw_data$"2017")
+data_2018 <- tibble(raw_data$"2018")
+data_2019 <- tibble(raw_data$"2019")
+data_2020 <- tibble(raw_data$"2020")
+data_2021 <- tibble(raw_data$"2021")
+data_2022 <- tibble(raw_data$"2022")
 
-# As the downloaded data is in "xlsx" format and very large, a dataframe for each year's 
[…]

### 2024-11-26 scripts/03-clean_data.R
complete data cleaning and save analysis data

@@ -1,44 +1,76 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 6 April 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Cleans the raw Paramedic Services Incident Data from Open Data Toronto
+# Author: Denise Chang
+# Date: 25 November 2024
+# Contact: dede.chang@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: 02-download_data.R
 
 #### Workspace setup ####
 library(tidyverse)
+library(openxlsx)
+library(janitor)
+library(arrow)
 
 #### Clean data ####
-raw_data <- read_csv("inputs/data/plane_data.csv")
 
-cleaned_data <-
-  raw_data |>
-  janitor::clean_names() |>
-  select(wing_width_mm, wing_length_mm, flying_time_sec_first_timer) |>
-  filter(wing_width_mm != "caw") |>
+# read the different XLSX files
+data_2017 <-
+  read_xlsx("data/01-raw_data/paramedic_services_2017.xlsx")
+data_2018 <-
+  read_xlsx("data/01-raw_data/paramedic_services_2018.xlsx")
+data_2019 <-
+  read_xlsx("data/01-raw_data/paramedic_services_2019.xlsx")
+data_2020 <-
+  read_xlsx("data/01-raw_data/paramedic_services_2020.xlsx")
+data_2021 <-
+  read_xlsx("data/01-raw_data/paramedic_services_2021.xlsx")
+data_2022 <-
+  read_xlsx("data/01-raw_data/paramedic_services_2022.xlsx")
+
+# combine files into one dataframe
+all_data <- bind_rows(data_2017,
+                      data_2018,
+                      data_2019,
+                      data_2020,
+                      data_2021,
+                      data_2022)
+
+expanded_data <-
+  clean_names(all_data) |>
+  # remove the entries with missing values
+  filter(!is.na(dispatch_time),!is.na(incident_type)) |>
+  # make incident_type uniform
+  mutate(incident_type = tolower(incident_type)) |>
+  filter(incident_type != "-", incident_type != "airport standby") |>
[…]

### 2024-11-26 scripts/03-clean_data.R
fix 'year' column type

@@ -45,7 +45,7 @@ expanded_data <-
   filter(incident_type != "-", incident_type != "airport standby") |>
   # separate time factors from dispatch time
   mutate(
-    year = year(dispatch_time),
+    year = as.integer(year(dispatch_time)),
     month = month(dispatch_time, label = TRUE),
     day_of_week = wday(dispatch_time, label = TRUE),
     hour = hour(dispatch_time)

### 2024-11-26 scripts/04-test_analysis_data.R
complete test script

@@ -1,69 +1,94 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 26 September 2024 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Tests the clean Paramedic Services Incident data
+# Author: Denise Chang
+# Date: 25 November 2024
+# Contact: dede.chang@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
+# Pre-requisites: 
+# - 02-download_data.R
+# - 03-clean_data.R
 
 #### Workspace setup ####
-library(tidyverse)
 library(testthat)
+library(arrow)
+library(tidyverse)
 
-data <- read_csv("data/02-analysis_data/analysis_data.csv")
-
+clean_data <- read_parquet("data/02-analysis_data/analysis_data.parquet")
 
 #### Test data ####
-# Test that the dataset has 151 rows - there are 151 divisions in Australia
-test_that("dataset has 151 rows", {
-  expect_equal(nrow(analysis_data), 151)
+
+## Test that there are no missing data
+# Test that there are no missing values in critical columns
+test_that("no missing values in critical columns", {
+  expect_true(all(!is.na(clean_data$year)))
+  expect_true(all(!is.na(clean_data$incident_type)))
+  expect_true(all(!is.na(clean_data$units_arrived_at_scene)))
 })
 
-# Test that the dataset has 3 columns
-test_that("dataset has 3 columns", {
-  expect_equal(ncol(analysis_data), 3)
+# Test that 'incident_type' does not contain "-"
+test_that("'incident_type' does not contain '-'", {
+  expect_false("-" %in% clean_data$incident_type)
 })
 
-# Test that the 'division' column is character type
-test_that("'division' is character", {
-  expect_type(analysis_data$division, "character")
+# Test that there are no empty strings in critical columns
+test_that("no empty strings in critical columns", {
+  expect_false(any(clean_data$avg_units_arrived == 0 | 
+                     clean_data$count == 0))
 })
 
-# Test that the 'party' column is character type
-test_that("'party' is character", {
-  expect_type(analysis_data$party, "character")
+## Test that the variables are of appropriate type
+# Test that the 'year' column is integer type
[…]

### 2024-11-26 scripts/03-clean_data.R
remove whitespace in "incident_type" column and fix path to read parquet in test script

@@ -41,7 +41,7 @@ expanded_data <-
   # remove the entries with missing values
   filter(!is.na(dispatch_time),!is.na(incident_type)) |>
   # make incident_type uniform
-  mutate(incident_type = tolower(incident_type)) |>
+  mutate(incident_type = trimws(tolower(incident_type))) |>
   filter(incident_type != "-", incident_type != "airport standby") |>
   # separate time factors from dispatch time
   mutate(

### 2024-11-26 scripts/04-test_analysis_data.R

@@ -11,9 +11,9 @@
 #### Workspace setup ####
 library(testthat)
 library(arrow)
-library(tidyverse)
+library(here)
 
-clean_data <- read_parquet("data/02-analysis_data/analysis_data.parquet")
+clean_data <- read_parquet(here("data/02-analysis_data/analysis_data.parquet"))
 
 #### Test data ####
 
@@ -22,7 +22,7 @@ clean_data <- read_parquet("data/02-analysis_data/analysis_data.parquet")
 test_that("no missing values in critical columns", {
   expect_true(all(!is.na(clean_data$year)))
   expect_true(all(!is.na(clean_data$incident_type)))
-  expect_true(all(!is.na(clean_data$units_arrived_at_scene)))
+  expect_true(all(!is.na(clean_data$avg_units_arrived)))
 })
 
 # Test that 'incident_type' does not contain "-"
@@ -32,8 +32,8 @@ test_that("'incident_type' does not contain '-'", {
 
 # Test that there are no empty strings in critical columns
 test_that("no empty strings in critical columns", {
-  expect_false(any(clean_data$avg_units_arrived == 0 | 
-                     clean_data$count == 0))
+  expect_false(any(clean_data$avg_units_arrived == "" | 
+                     clean_data$count == ""))
 })
 
 ## Test that the variables are of appropriate type
@@ -59,7 +59,7 @@ test_that("'hour' is integer", {
 
 # Test that the 'incident_type' column is character type
 test_that("'incident_type' is character", {
-  expect_type(clean_data$hour, "character")
+  expect_type(clean_data$incident_type, "character")
 })
 
 # Test that the 'avg_units_arrived' column is double type
@@ -84,8 +84,9 @@ test_that("'hour' is between 0 and 23", {
 })
 
 # Test that 'incident_type' contains only valid values
+valid_types <- c("emergency transfer", "fire", "medical", "motor vehicle accident")
 test_that("'incident_type' contains only valid values", {
-  expect_true(all(data$incident_type %in% c("emergency transfer", "fire", "medical", "motor vehicle accident")))
+  expect_true(all(clean_data$incident_type %in% valid_types))
 })
 
 # Test that 'avg_units_arrived' column values are non-negative

### 2024-11-26 scripts/00-simulate_data.R
complete simulation script (00) and use code styler (04)

@@ -1,52 +1,59 @@
 #### Preamble ####
-# Purpose: Simulates a dataset of Australian electoral divisions, including the 
-  #state and party that won each division.
-# Author: Rohan Alexander
-# Date: 26 September 2024
-# Contact: rohan.alexander@utoronto.ca
+# Purpose: Simulates a dataset of paramedics incidents in Toronto from 2017 to 2022
+# Author: Denise Chang
+# Date: 26 November 2024
+# Contact: dede.chang@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: The `tidyverse` package must be installed
-# Any other information needed? Make sure you are in the `starter_folder` rproj
-
+# Pre-requisites: None
 
 #### Workspace setup ####
 library(tidyverse)
-set.seed(853)
+library(arrow)
 
+set.seed(304) # random seed for reproducibility
 
-#### Simulate data ####
-# State names
-states <- c(
-  "New South Wales",
-  "Victoria",
-  "Queensland",
-  "South Australia",
-  "Western Australia",
-  "Tasmania",
-  "Northern Territory",
-  "Australian Capital Territory"
+#### Simulate Data ####
+incident_types <-
+  c("emergency transfer",
+    "fire",
+    "medical",
+    "motor vehicle accident")
+months <- month.name  # Full month names
+days_of_week <- c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")
+
+# Create all combinations of year, month, and day_of_week (subset of real dataset)
+year_month_day_combinations <- expand.grid(
+  year = sample(2017:2022, 10, replace = TRUE),
+  month = sample(months, 10, replace = TRUE),
+  day_of_week = sample(days_of_week, 10, replace = TRUE)
 )
 
-# Political parties
-parties <- c("Labor", "Liberal", "Greens", "National", "Other")
+# Create ordered factors for month and day_of_week
+year_month_day_combinations$month <-
+  factor(year_month_day_combinations$month,
+         levels = months,
+         ordered = TRUE)
+year_month_day_combinations$day_of_week <-
+  factor(year_month_day_combinations$day_of_week,
+         levels = days_of_week,
+         ordered = TRUE)
 
-# Create a dataset by randomly assigning states and parties to divisions
-analysis_data <- tibble(
-  division = paste("Division", 1:151),  # Add "Division" to make it a character
-  state = sample(
-    states,
-    size = 151,
-    replace = TRUE,
[…]

### 2024-11-26 scripts/04-test_analysis_data.R

@@ -4,7 +4,7 @@
 # Date: 25 November 2024
 # Contact: dede.chang@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: 
+# Pre-requisites:
 # - 02-download_data.R
 # - 03-clean_data.R
 
@@ -13,7 +13,8 @@ library(testthat)
 library(arrow)
 library(here)
 
-clean_data <- read_parquet(here("data/02-analysis_data/analysis_data.parquet"))
+clean_data <-
+  read_parquet(here("data/02-analysis_data/analysis_data.parquet"))
 
 #### Test data ####
 
@@ -32,7 +33,7 @@ test_that("'incident_type' does not contain '-'", {
 
 # Test that there are no empty strings in critical columns
 test_that("no empty strings in critical columns", {
-  expect_false(any(clean_data$avg_units_arrived == "" | 
+  expect_false(any(clean_data$avg_units_arrived == "" |
                      clean_data$count == ""))
 })
 
@@ -84,7 +85,11 @@ test_that("'hour' is between 0 and 23", {
 })
 
 # Test that 'incident_type' contains only valid values
-valid_types <- c("emergency transfer", "fire", "medical", "motor vehicle accident")
+valid_types <-
+  c("emergency transfer",
+    "fire",
+    "medical",
+    "motor vehicle accident")
 test_that("'incident_type' contains only valid values", {
   expect_true(all(clean_data$incident_type %in% valid_types))
 })

### 2024-11-27 paper/paper.qmd
complete introduction and add data section ideas. Update references.bib accordingly

@@
   - Denise Chang
-thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
+thanks: "Code and data are available at: [https://github.com/DeniseChang9/Emergency_service_demands.git](https://github.com/DeniseChang9/Emergency_service_demands.git)."
 date: today
@@
 #| message: false
-#| eval: false
 
@@
 
-Overview paragraph
+"In an emergency, seconds count." [@city_of_toronto_public_safety_alerts].
+Emergency medical services (EMS) are essential to public health by providing care during life-threatening situations.
+However, Toronto’s paramedic services are struggling to meet demand.
+Toronto Auditor General’s Office reports that there were over 1,200 episodes in 2023 where no ambulances were available.
+[@toronto_auditor_paramedic_services].
+To echo this report, this paper examines EMS demand using Toronto Paramedic Services’ incident data from Open Data Toronto.
 
-Estimand paragraph
+In this analysis, I am interested in EMS demand against time-based factors in Toronto from 2017 to 2022.
+I explore trends in paramedic services, such as dispatch time, incident type, and number of units dispatched per incident to identify periods of high-volume demand and low availability of resources.
+I find that ...
+[ADD A FEW RESULTS HERE (or in a different paragraph??)]
 
-Results paragraph
-
-Why it matters paragraph
-
-Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....
+The remainder of this paper is organized as follows.
+@sec-data discusses the data source and an overview of the studied variables.
+@sec-model constructs a model that predicts a shortage in paramedic resources based on time factors.
+@sec-results presents the results and findings of the exploration of the data.
+@sec-discussion discusses implications, limitations, and suggestions for future research.
+[TELEGRAPH APPENDIX HERE]
 
@@
 
+what data was used (from Open Data Toronto) + name and mention last dat accessed. What frequency it's updated etc etc
+which libraries were used
+
 We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...
 
-Overview text
+## Measurements
 
-## Measurement
+The dataset captures ambulance incident records with real-world emergency service activities.
 
-Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.
+[INSERT SAMPLE OF CLEAN DATASET]
 
-## Outcome variables
+Table 1 is a sample of the clean dataset used in this paper.
+Each row in the dataset represents to a different incident.
+More information such as the type of emergency and paramedic response are also included and are unique to the incident.
 
-Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.
+The temporal variables (`year`, `month`, `day_of_week`, and `hour`) are derived from dispatch timestamp recorded during each incident.
+These entries useful in exploring seasonal, weekly, and daily patterns.
 
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
+The incident type (`incident_type`) represents a classification provided by dispatchers or paramedics based on the nature of the call.
+This variable reflects categories such as medical emergencies, vehicle accidents, and fires.
+It ensures that diverse incidents are grouped into manageable classifications for analysis.
 
-```{r}
-#| label: fig-bills
-#| fig-cap: Bills of penguins
-#| echo: false
-#| eval: false
+The response metrics (`avg_units_arrived` and `count`) represent operational data tied to each incident.
+The `avg_units_arrived` variable is calculated by averaging the number of units dispatched to each type of incident within specific time intervals.
+The `count` variable represents the number of incidents recorded for each combination of temporal and incident characteristics.
+These variables quantify resource utilization and demand, enabling a detailed exploration of service patterns.
 
-ggplot(penguins, aes(x = island, fill = species)) +
-  geom_bar(alpha = 0.8) +
-  scale_fill_manual(values = c("darkorange","purple","cyan4"),
-                    guide = "none") +
-  theme_minimal() +
-  facet_wrap(~species, ncol = 1) +
-  coord_flip()
-```
+Through this systematic representation of real-world phenomena, the dataset provides a foundation for analyzing ambulance demand and the factors influencing emergency service delivery.
 
-Talk more about it.
+## Outcome Variables
 
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
+This analysis focuses on two outcome variables that represent ambulance response activity:
 
-```{r}
-#| label: fig-planes
-#| fig-cap: Relationship between wing length and width
-#| echo: false
-#| warning: false
-#| message: false
-#| eval: false
+### `avg_units_arrived`
+This variable measures the average number of ambulance units arriving at incidents. It is continuous and provides insight into the resource intensity of different types of calls.
 
-analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))
+- **Visualization**: A violin plot is used to examine the distribution of `avg_units_arrived` across different `incident_type` values.
+Temporal trends are visualized with a line chart showing annual changes in `avg_units_arrived`.
 
-analysis_data |> 
-  ggplot(aes(x = width, y = length)) +
-  geom_point(alpha = 0.8) +
-  theme_minimal() +
-  labs(x = "Wing width (mm)",
-       y = "Wing length (mm)")
-```
+### `count`
+This variable captures the number of incidents for each category and is integer-valued. It reflects the frequency of calls and is critical for understanding demand patterns.
 
-Talk way more about it.
+- **Visualization**: Bar charts display the total counts of incidents by `incident_type`, while a heatmap reveals temporal patterns by plotting `count` across combinations of `hour`, `day_of_week`, and `month`.
 
-## Predictor variables
+## Predictor Variables
 
-Add graphs, tables and text.
+Several variables are explored as predictors to analyze temporal, categorical, and spatial factors influencing ambulance response:
 
-Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.
+### Temporal Predictors
+- **`year`**: Indicates the year of each incident, spanning 2017 to 2022.  
+  - **Visualization**: A line chart visualizes changes in `count` and `avg_units_arrived` over time to detect trends.
 
-# Model
+- **`month`**: An ordered factor capturing the month of the incident.  
+  - **Visualization**: A seasonal heatmap overlays `month` and `day_of_week` to analyze patterns in incident frequency and average unit arrival.
+
+- **`day_of_week`**: An ordered factor representing the day of the week.  
+  - **Visualization**: Temporal variability is further explored with bar charts and heatmaps.
+
+- **`hour`**: Represents the hour of the day (0–23).  
+  - **Visualization**: A density plot of incident counts by hour reveals peak times for service demands.
+
+### Categorical Predictor
+- **`incident_type`**: A categorical variable indicating the type of incident (e.g., medical emergencies, vehicle accidents).  
+  - **Visualization**: Stacked bar charts and facet grids illustrate differences in resource use and frequency across `incident_type`.
+
+## Summary Statistics and Relationships
+
+To understand the dataset, summary statistics are computed for `avg_units_arrived` and `count`, including mean, median, and standard deviation.
+Visualizations such as histograms and scatter plots capture distributions and relationships between the variables.
+A scatter plot matrix examines pairwise relationships, while a correlation heatmap identifies patterns among numerical variables.
+
+Together, these outcomes and predictors provide a detailed view of ambulance demand and the factors influencing it.
+
+# Model {#sec-model}
 
@@
 
-# Results
+# Results {#sec-results}
 
@@
 
-# Discussion
+# Discussion {#sec-discussion}
 

### 2024-12-02 paper/paper.qmd
complete data overview and data source with references

@@
 
-## Overview
+The statistical programming language R [@R] is used to process, manage and visualize the data.
+Specifically, statistical libraries such as `opendatatoronto` [@opendatatoronto], `openxlsx` [@openxlsx] and `janitor` [@janitor] are used to simulate, download and clean the raw data.
+The library `arrow` [@arrow] was used to save the simulated dataset and the analysis dataset.
+Other libraries like '`knitr` [@knitr], `here` [@here] are used to load and to render tables.
+The library `tidyverse` [@tidyverse] is useful throughout the entire data manipulation process.
 
-what data was used (from Open Data Toronto) + name and mention last dat accessed. What frequency it's updated etc etc
-which libraries were used
+## Data source
 
-We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...
+The dataset used for this analysis is titled "Paramedic Services Incident Data" and is published by Toronto Paramedic Services [@data_parmedic_source].
+For this paper, the dataset is retrieved from the City of Toronto Open Data Portal through the R package `opendatatoronto` [@opendatatoronto].
+The initial dataset features paramedic dispatch time, the type of incident, the priority level of each incident, the number of paramedic units arrived at scene and the forward station area of the incident.
+These features are annually refreshed on Open Data Toronto by Toronto Paramedic Services.
+
+The data used for this paper was retrieved on November 25, 2024 and was last refreshed on October 5, 2023.
 

### 2024-12-03 paper/paper.qmd
fix measurement in data

@@
 library(tidyverse)
-library(palmerpenguins)
+library(openxlsx)
+library(readxl)
+library(arrow)
+library(here)
+library(knitr)
+
 ```
@@
 
+## Overview
+The dataset used for this analysis is titled "Paramedic Services Incident Data" and is published by Toronto Paramedic Services [@data_parmedic_source].
+For this paper, the dataset is retrieved from the City of Toronto Open Data Portal.
+
 The statistical programming language R [@R] is used to process, manage and visualize the data.
 Specifically, statistical libraries such as `opendatatoronto` [@opendatatoronto], `openxlsx` [@openxlsx] and `janitor` [@janitor] are used to simulate, download and clean the raw data.
-The library `arrow` [@arrow] was used to save the simulated dataset and the analysis dataset.
-Other libraries like '`knitr` [@knitr], `here` [@here] are used to load and to render tables.
+Libraries like `arrow` [@arrow] and `readxl` [@readxl] were used to save and read datasets.
+Other libraries like `knitr` [@knitr], `here` [@here] are used to load and to render tables.
 The library `tidyverse` [@tidyverse] is useful throughout the entire data manipulation process.
 
-## Data source
-
-The dataset used for this analysis is titled "Paramedic Services Incident Data" and is published by Toronto Paramedic Services [@data_parmedic_source].
-For this paper, the dataset is retrieved from the City of Toronto Open Data Portal through the R package `opendatatoronto` [@opendatatoronto].
-The initial dataset features paramedic dispatch time, the type of incident, the priority level of each incident, the number of paramedic units arrived at scene and the forward station area of the incident.
+The initial dataset features data on paramedic dispatch time, type of incident, priority level of each incident, number of paramedic units arrived at scene and forward sortation area of the incident.
 These features are annually refreshed on Open Data Toronto by Toronto Paramedic Services.
@@
 
-The dataset captures ambulance incident records with real-world emergency service activities.
+The incident data is saved in different files according to the year of incident. The variables and measurement method of the variables are preserved throughout the recorded years.
 
-[INSERT SAMPLE OF CLEAN DATASET]
+```{r}
+#| message: false
+#| echo: false
+#| warning: false
+#| label: tbl-raw_data
+#| tbl-cap: First Ten Rows of the Paramedic Services Incident Data from 2017
 
-Table 1 is a sample of the clean dataset used in this paper.
-Each row in the dataset represents to a different incident.
-More information such as the type of emergency and paramedic response are also included and are unique to the incident.
+# read raw data
+raw_2017 <- read_xlsx(here::here("data/01-raw_data/paramedic_services_2017.xlsx"))
 
-The temporal variables (`year`, `month`, `day_of_week`, and `hour`) are derived from dispatch timestamp recorded during each incident.
-These entries useful in exploring seasonal, weekly, and daily patterns.
+# visualize the first 10 rows of the data
+raw_2017 |>
+  select(Dispatch_Time, Incident_Type, Priority_Number, Units_Arrived_At_Scene, Forward_Sortation_Area) |>
+  slice(1:10) |>
+  kable(col.names = c("Dispatch Time", "Incident Type", "Priority Number", "Units Arrived", "Location"))
+```
 
-The incident type (`incident_type`) represents a classification provided by dispatchers or paramedics based on the nature of the call.
-This variable reflects categories such as medical emergencies, vehicle accidents, and fires.
-It ensures that diverse incidents are grouped into manageable classifications for analysis.
-
-The response metrics (`avg_units_arrived` and `count`) represent operational data tied to each incident.
-The `avg_units_arrived` variable is calculated by averaging the number of units dispatched to each type of incident within specific time intervals.
-The `count` variable represents the number of incidents recorded for each combination of temporal and incident characteristics.
-These variables quantify resource utilization and demand, enabling a detailed exploration of service patterns.
-
-Through this systematic representation of real-world phenomena, the dataset provides a foundation for analyzing ambulance demand and the factors influencing emergency service delivery.
+@tbl-raw_data is a sample of the 10 first entries in the incident data in 2017.
+Each row in the dataset represents a unique incident.
+"Dispatch Time" is the precise time when the first paramedic unit was assigned to an incident.
+This is measured in year, month, date, hour, minutes, and seconds.
+"Incident Type" is the category assigned to the incident by the dispatcher based on the information provided in the 9.1.1.
+call.
+The possible categories of incident are medical emergencies, emergency transfers, fire-related incidents and motor vehicle accidents.
+"Priority number" represents the urgency of an incident with 1 being the most urgent and 5 being the least.
+The priority number is measured by the Medical Priority Dispatch System (MPDS) based on the information provided by the 9.1.1.
+caller(s).
+"Units Arrived" is the total number of paramedics that arrived on the scene of incident.
+This is measured by counting the number of different paramedic units who were dispatched and responded to the incident.
+Units part of this count do not have to be simultaneously present, and a unit who leaves the scene and comes back afterwards is considered 1 count.
+"Location" is the general location of the incident based on Postal Code Forward Sortation Areas. This is determined as the first three characters of the postal code of the incident location.
 

### 2024-12-03 scripts/03-clean_data.R
Adjust decimals in avg units

@@ -66,7 +66,7 @@ clean_data <-
   selected_data |>
   group_by(year, month, day_of_week, hour, incident_type) |>
   summarise(
-    avg_units_arrived = mean(units_arrived_at_scene, na.rm = TRUE),
+    avg_units_arrived = format(round(mean(units_arrived_at_scene, na.rm = TRUE), 1), nsmall = 1),
     count = n()
   ) |>
   ungroup()

### 2024-12-03 scripts/03-clean_data.R
make avg units type dbl

@@ -66,7 +66,7 @@ clean_data <-
   selected_data |>
   group_by(year, month, day_of_week, hour, incident_type) |>
   summarise(
-    avg_units_arrived = format(round(mean(units_arrived_at_scene, na.rm = TRUE), 1), nsmall = 1),
+    avg_units_arrived = as.numeric(format(round(mean(units_arrived_at_scene, na.rm = TRUE), 1), nsmall = 1)),
     count = n()
   ) |>
   ungroup()

### 2024-12-03 scripts/05-model_data.R
builds model in script

@@ -1,37 +1,45 @@
 #### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Models number of calls based on temporal factors, incident type and average units on incident scene
+# Author: Denise Chang
+# Date: 2 December 2024
+# Contact: dede.chang@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
+# Pre-requisites: 02-download_data.R and 03-clean_data.R
 
 #### Workspace setup ####
 library(tidyverse)
 library(rstanarm)
+library(arrow)
 
-#### Read data ####
-analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
+#### Model data ####
+# Read data
+clean_data <-
+  read_parquet("data/02-analysis_data/analysis_data.parquet")
 
-### Model data ####
-first_model <-
-  stan_glm(
-    formula = flying_time ~ length + width,
-    data = analysis_data,
-    family = gaussian(),
-    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_aux = exponential(rate = 1, autoscale = TRUE),
-    seed = 853
+# Prepare data
+clean_data <- clean_data |>
+  mutate(
+    avg_units_arrived = as.numeric(avg_units_arrived),
+    month = factor(month, levels = month.abb, ordered = TRUE),
+    day_of_week = factor(
+      day_of_week,
+      levels = c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat"),
+      ordered = TRUE
+    ),
+    incident_type = factor(incident_type)
   )
 
-
-#### Save model ####
-saveRDS(
-  first_model,
-  file = "models/first_model.rds"
+# Fit the model using rstanarm
+incident_model <- stan_glm(
+  formula = count ~ year + month + day_of_week + hour + incident_type + avg_units_arrived,
+  data = clean_data,
+  family = gaussian(),
+  prior = normal(0, 2.5),
+  prior_intercept = normal(0, 2.5),
+  prior_aux = exponential(1),
[…]

### 2024-12-03 paper/paper.qmd
complete data section and model section.

@@
 library(knitr)
-
+library(modelsummary)
+library(patchwork)
 ```
@@
 This is measured in year, month, date, hour, minutes, and seconds.
-"Incident Type" is the category assigned to the incident by the dispatcher based on the information provided in the 9.1.1.
-call.
+"Incident Type" is the category assigned to the incident by the dispatcher based on the information provided by the 9.1.1.
+caller(s).
 The possible categories of incident are medical emergencies, emergency transfers, fire-related incidents and motor vehicle accidents.
@@
 
-## Outcome Variables
+## Variables of Interest
 
-This analysis focuses on two outcome variables that represent ambulance response activity:
+In the analysis dataset, most variables were constructed using the variables in the raw dataset.
+From the original dataset, "Dispatch Time" was broken down into four separate variables in the analysis dataset: "Year", "Month", "Day of the Week" and "Hour of the Day".
+The aim for this construction is to isolate the effect of each temporal factor on the paramedic incident request paramedic response.
+While holding the time variables and the "Incident Type" variable constant, two more variables were constructed from the "Units Arrived" variable from the raw dataset: "Average Units Arrived" and "Number of Incidents".
 
-### `avg_units_arrived`
-This variable measures the average number of ambulance units arriving at incidents. It is continuous and provides insight into the resource intensity of different types of calls.
+```{r}
+#| message: false
+#| echo: false
+#| warning: false
+#| label: tbl-avg_units
+#| tbl-cap: Ten First Rows of the Analysis Data on Paramedic Services Incident Data from 2017 to 2022
 
-- **Visualization**: A violin plot is used to examine the distribution of `avg_units_arrived` across different `incident_type` values.
-Temporal trends are visualized with a line chart showing annual changes in `avg_units_arrived`.
+# read raw data
+clean_data <- read_parquet(here::here("data/02-analysis_data/analysis_data.parquet"))
 
-### `count`
-This variable captures the number of incidents for each category and is integer-valued. It reflects the frequency of calls and is critical for understanding demand patterns.
+# visualize the first 10 rows of the data
+clean_data |>
+  slice(1:10) |>
+  kable(col.names = c("Year", "Month", "Day of the Week", "Hour of the Day", "Incident Type", "Average Units Arrived", "Number of Incidents" ))
 
-- **Visualization**: Bar charts display the total counts of incidents by `incident_type`, while a heatmap reveals temporal patterns by plotting `count` across combinations of `hour`, `day_of_week`, and `month`.
+```
+@tbl-avg_units is a sample of the analysis dataset used in this paper. The variables of interest of the analysis dataset are:
 
-## Predictor Variables
+- Number of incidents:This variable represents the total count of emergency incidents recorded for each unique combination of time and incident-type factors.
+This is the outcome variable in the model, since it reflects the demandfor paramedic services.
 
-Several variables are explored as predictors to analyze temporal, categorical, and spatial factors influencing ambulance response:
+- Average number of units arrived: This variable reflects the average number of paramedic units dispatched and arrived at the scene for a given combination of incident type, time, and date.
+This is another outcome variable , since it provides insight into resource allocation.
 
-### Temporal Predictors
-- **`year`**: Indicates the year of each incident, spanning 2017 to 2022.  
-  - **Visualization**: A line chart visualizes changes in `count` and `avg_units_arrived` over time to detect trends.
+- Incident Type: This categorical variable describes the nature of each emergency, such as medical emergencies, fire-related incidents, motor vehicle accidents, or emergency transfers.
+It captures the diversity of incidents and is useful for exploring differences in ressource needs
 
-- **`month`**: An ordered factor capturing the month of the incident.  
-  - **Visualization**: A seasonal heatmap overlays `month` and `day_of_week` to analyze patterns in incident frequency and average unit arrival.
+- Year: The year the incident occurred, capturing long-term trends in incident frequency.
 
-- **`day_of_week`**: An ordered factor representing the day of the week.  
-  - **Visualization**: Temporal variability is further explored with bar charts and heatmaps.
+- Month: The month of occurrence, which higlights seasonal trends
 
-- **`hour`**: Represents the hour of the day (0–23).  
-  - **Visualization**: A density plot of incident counts by hour reveals peak times for service demands.
+- Day of Week: The day on which the incident took place, accounting for patterns related to weekdays versus weekends.
 
-### Categorical Predictor
-- **`incident_type`**: A categorical variable indicating the type of incident (e.g., medical emergencies, vehicle accidents).  
-  - **Visualization**: Stacked bar charts and facet grids illustrate differences in resource use and frequency across `incident_type`.
-
-## Summary Statistics and Relationships
-
-To understand the dataset, summary statistics are computed for `avg_units_arrived` and `count`, including mean, median, and standard deviation.
-Visualizations such as histograms and scatter plots capture distributions and relationships between the variables.
-A scatter plot matrix examines pairwise relationships, while a correlation heatmap identifies patterns among numerical variables.
-
-Together, these outcomes and predictors provide a detailed view of ambulance demand and the factors influencing it.
+- Hour: The hour of the day, capturing diurnal patterns in emergency service demand.
 
@@
 ## Model set-up
+We run the model in R [@R] using the `rstanarm` package [@rstanarm]. Default priors from `rstanarm` are applied.
+The estimating equating is as follows:
 
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
+Define $y_i$ as the number of incidents observed for a specific combination of factors. Then, $\mu_i$ is the expected number of incidents based on temporal and incident-specific predictors.
 
 \begin{align}  
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
+y_i | \mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\  
+\mu_i &= \alpha + \beta_{\text{year}} \cdot \text{year}_i + \sum_{k=1}^{11} \beta_{\text{month}, k} \cdot \text{month}_{ik} + \sum_{j=1}^{6} \beta_{\text{day}, j} \cdot \text{day}_{ij} \\
+&\quad + \beta_{\text{hour}} \cdot \text{hour}_i + \sum_{t=1}^{3} \beta_{\text{incident}, t} \cdot \text{incident}_{it} + \beta_{\text{units}} \cdot \text{avg\_units}_i \\  
 \alpha &\sim \mbox{Normal}(0, 2.5) \\  
 \beta &\sim \mbox{Normal}(0, 2.5) \\  
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
 \sigma &\sim \mbox{Exponential}(1)  
 \end{align}  
+In this model, we define the variables as follows:
 
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
+- $y_i$ is the number of incidents observed for a specific combination of factors
+
+- $\mu_i$ is the expected number of incidents based on predictors
+
+- $\alpha$ is the intercept term, representing the baseline number of incidents.
+
+- $\beta_{\text{year}}$ is the coefficient for the year variable.
+
+- $\beta_{\text{month}}$ is the set of coefficients for the month of the year.
+
+- $\beta_{\text{day}}$ is the set of coefficients for the day of the week.
+
+- $\beta_{\text{hour}}$ is the coefficient for the hour of the day.
+
+- $\beta_{\text{incident}}$ is the set of coefficients for the incident types.
+
+- $\beta_{\text{units}}$ is the coefficient for the average number of units arriving on scene
+
+- $\sigma$ is the standard deviation of the residuals, representing unexplained variability in incident counts.
 
 ### Model justification
+For this analysis, a Bayesian Generalized Linear Model (GLM) was chosen to predict incident call counts as a function of temporal factors, incident type, and resource allocation.
+This approach provides the flexibility to model log-transformed continuous outcomes and includes prior beliefs about the relationships between predictors and the outcome.
+The Bayesian framework quantifies uncertainty in parameter estimates and predictions, which is useful in our context where variability and uncertainty are prevalent.
+This model is appropriate given the observed overdispersion in the count data.
 
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
+In Bayesian modeling, priors were defined to reflect plausible values.
+A $\mbox{Normal}(0, 10)$ prior was used for the intercept ($\alpha$) to allow a wide range of baseline incident counts across predictor combinations.
+Predictor coefficients ($\beta$) were assigned a $\mbox{Normal}(0, 2.5)$ prior to accommodate moderate effect sizes without being too restrictive.
+The residual standard deviation ($\sigma$) was given an $\mbox{Exponential}(1)$ prior to emphasize smaller residual errors.
+These priors avoid constraints that are too narrow, reducing bias in the estimates
 
-We can use maths by including latex between dollar signs, for instance $\theta$.
+This model relies on several assumptions.
+First, it assumes a linear relationship between predictors and the log-transformed outcome variable, which simplifies interpretation.
+The model also assumes residuals follow a Gaussian distribution, consistent with the log-transformation of the outcome variable.
+Despite its strengths, the model has limitations.
+It does not explicitly model nonlinear relationships or interactions,like those between temporal and incident type variables, which could capture more complex dynamics.
+Finally, the Gaussian residual assumption may be inappropriate in certain cases of extreme overdispersion.
+
+Several alternative models were considered.
+Poisson regression, a natural choice for count data, was notr chosen due to overdispersion in the outcome variable.
+Negative binomial regression was also considered, but does not incorporate prior beliefs or uncertainties as naturally as Bayesian methods.
+
+Ultimately, the Bayesian Gaussian GLM was selected since it can handle continuous log-transformed data and include prior beliefs while still being digesteable.Through Bayesian inference, the model quantifies uncertainty and provides a probabilistic framework ideal for emergency response planning.
+This approach ensures actionable predictions for resource allocation decisions.
 
 # Results {#sec-results}
-
-Our results are summarized in @tbl-modelresults.
+Our results are summarized in @tbl-model.
 
 ```{r}
+#| message: false
+#| warning: false
 #| echo: false
-#| eval: false
-#| warning: false
-#| message: false
+#| label: tbl-model
+#| tbl-cap: Model Summary of Incident Count
 
-library(rstanarm)
+# read model
+model <-
+  readRDS(file = here::here("models/incident_model.rds"))
 
-first_model <-
-  readRDS(file = here::here("models/first_model.rds"))
+summary(model)
 ```
 
+The Bayesian GLM model, based on 38,494 observations and 24 predictors, has a posterior predictive mean (mean_PPD) of 44.4, which aligns with observed incident counts.
+This shows that the model captures variability.
+
+Temporal factors play a role in determining incident volume.
+The coefficient for hour (mean = 1.2) shows predictable variations in demand throughout the day, with highly precise estimates (sd = 0.0).
+Monthly and daily trends are captured through polynomial terms, reflecting the nonlinear nature of incident occurrences across different times of the year and week.
+
+The analysis suggests the dominance of medical incidents in driving overall incident counts.
+The coefficient for medical calls (mean = 126.4) is much higher than the coefficient for fire-related incidents (mean = -6.6) and the coefficient for motor vehicle accidents (mean = 0.2).
+This shows that medical emergencies is the primary source of paramedic service demand, requiring special attention in operational planning.
+
+The positive relationship between the average number of units arriving and incident counts (mean = 1.3) suggests that resource allocation is another determining factor.
+There is a correlation where increased resource availability may correspond to higher incident volumes.
+This reflects areas of concentrated demand and hints at resource allocation strategies.
+
+## Hour of the Day on Paramedic Incident Demand and Reponse
 ```{r}
+#| message: false
+#| warning: false
 #| echo: false
-#| eval: false
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
+#| label: tbl-clock_plot
+#| tbl-cap: Paramedic Incident Counts and Number of Units Arrived on Scene by Hour of the Day.
 
-modelsummary::modelsummary(
-  list(
-    "First model" = first_model
-  ),
-  statistic = "mad",
-  fmt = 2
+# Filter relevant columns
+filtered_data <- clean_data %>%
+  select(hour, count, avg_units_arrived)
+
+# Summarize counts and average units by hour
+hourly_summary <- filtered_data %>%
+  group_by(hour) %>%
+  summarize(
+    total_count = sum(count),
+    avg_units = mean(avg_units_arrived)
   )
+
+# Clock plot for incident count
+count_plot <- ggplot(hourly_summary, aes(x = hour, y = total_count)) +
+  geom_col(fill = "steelblue", color = "black") +
+  scale_x_continuous(breaks = 0:23, limits = c(0, 23)) +
+  coord_polar(start = 0) +
+  theme_minimal() +
+  labs(
+    title = "Incident Counts by Hour",
+    x = "Hour (Clock Position)",
+    y = "Number of Incidents"
+  )
+
+# Clock plot for average units arrived
+units_plot <- ggplot(hourly_summary, aes(x = hour, y = avg_units)) +
+  geom_col(fill = "coral", color = "black") +
+  scale_x_continuous(breaks = 0:23, limits = c(0, 23)) +
+  coord_polar(start = 0) +
+  theme_minimal() +
+  labs(
+    title = "Average Units Arrived by Hour",
+    x = "Hour (Clock Position)",
+    y = "Avg Units Arrived"
+  )
+
+# Combine the plots side by side
+combined_plot <- count_plot + units_plot
+
+# Display the plots
+invisible(print(combined_plot))
 ```

### 2024-12-03 paper/paper.qmd
complete paper

@@
 ---
-title: "My title"
-subtitle: "My subtitle if needed"
+title: "Understanding Paramedic Services Demand and Response"
+subtitle: "An Analysis of the Impact of Time-Based Factors on Paramedic Servie Demand in Toronto from 2017 to 2022"
 author: 
@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "This paper examines the trends paramedic service demand and response against time-based factors.
+Using the Paramedic Services Incident Data from 2017 to 2022,the analysis aims to determine the influence of temporal factors and incident types on paramedic service demand and response.
+The findings shows a constant paramedic response depite fluctuating service demands throughout seasons and time of the day suggesting a flaw in resource allocation.
+The results of this study are significant, as they can be indicators to policymakers and paramedic services during resource manadement and dispatch."
 format: pdf
@@
 library(patchwork)
+library(gridExtra)
 ```
@@
 However, Toronto’s paramedic services are struggling to meet demand.
-Toronto Auditor General’s Office reports that there were over 1,200 episodes in 2023 where no ambulances were available.
-[@toronto_auditor_paramedic_services].
+Toronto Auditor General’s Office reports that there were over 1,200 episodes in 2023 where no ambulances were available [@toronto_auditor_paramedic_services].
 To echo this report, this paper examines EMS demand using Toronto Paramedic Services’ incident data from Open Data Toronto.
@@
 I explore trends in paramedic services, such as dispatch time, incident type, and number of units dispatched per incident to identify periods of high-volume demand and low availability of resources.
-I find that ...
-[ADD A FEW RESULTS HERE (or in a different paragraph??)]
+I find that the number of incidents requiring paramedic services is higher in summer seasons, peaking in July and in August.
+The lowest number of incidents are in February.
+I also find that the number of incident requests are lowest between 1AM and 7AM, while the periods of high demand are between 11AM to 7PM.
+However, the number of paramedic units dispatched remained constant, despite the fluctuations in demand.
 
@@
 @sec-discussion discusses implications, limitations, and suggestions for future research.
-[TELEGRAPH APPENDIX HERE]
+[Appendix -@sec-observational_data] provides a detailed exploration of observational data collection methods.
 
@@
 Other libraries like `knitr` [@knitr], `here` [@here] are used to load and to render tables.
-The library `tidyverse` [@tidyverse] is useful throughout the entire data manipulation process.
+Libraries such as `gridExtra` [@gridExtra] and `patchwork` [@patchwork] are used in te visualization of the data.The library `rstanarm` [@rstanarm] is used in the modeling process and tidyverse` [@tidyverse] is useful throughout the entire data manipulation process.
 
@@
 
-The goal of our modelling strategy is twofold. Firstly,...
-
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
-
 ## Model set-up
@@
 # Results {#sec-results}
-Our results are summarized in @tbl-model.
+
+The results of the model can be found in @tbl-model.
 
@@
 
-# read model
-model <-
-  readRDS(file = here::here("models/incident_model.rds"))
+# Read the model
+model <- readRDS(file = here::here("models/incident_model.rds"))
 
-summary(model)
+# Extract coefficients from the model summary
+model_summary <- summary(model)
+
+# Convert the coefficients matrix to a data frame
+coefficients <- as.data.frame(model_summary)
+
+# Round the numeric columns to 2 decimal places
+coefficients[] <- lapply(coefficients, function(x) if(is.numeric(x)) round(x, 2) else x)
+
+# Round the last row to a smaller number of digits
+coefficients[nrow(coefficients),] <- format(coefficients[nrow(coefficients),], digits = 3)
+
+# Display the table without the 'Term' column
+kable(coefficients, caption = "Model Coefficients", format = "markdown")
+
 ```
@@
 ## Hour of the Day on Paramedic Incident Demand and Reponse
-```{r}
+```{r, fig.width=8, fig.height=4}
 #| message: false
@@
 #| echo: false
-#| label: tbl-clock_plot
-#| tbl-cap: Paramedic Incident Counts and Number of Units Arrived on Scene by Hour of the Day.
+#| label: fig-clock_plot
+#| fig-cap: Paramedic Incident Counts and Number of Units Arrived on Scene by Hour of the Day from 2017 to 2022
 
@@
   geom_col(fill = "steelblue", color = "black") +
-  scale_x_continuous(breaks = 0:23, limits = c(0, 23)) +
+  scale_x_continuous(breaks = 0:24, limits = c(0, 24)) +
   coord_polar(start = 0) +
@@
   geom_col(fill = "coral", color = "black") +
-  scale_x_continuous(breaks = 0:23, limits = c(0, 23)) +
+  scale_x_continuous(breaks = 0:24, limits = c(0, 24)) +
   coord_polar(start = 0) +
@@
     x = "Hour (Clock Position)",
-    y = "Avg Units Arrived"
+    y = "Average Units Arrived"
+  ) +
+  theme(
+    plot.margin = margin(20, 20, 20, 20), 
+    plot.title = element_text(hjust = 0.5)
   )
@@
 ```
+@fig-clock_plot shows distinct patterns in incident demand and resource allocation throughout the day.
+The number of incident calls is lowest during the early morning hours, from 1 AM to 7 AM.
+This suggests that fewer emergencies or incidents occur during these hours, possibly due to lower activity levels at night.
+Conversely, from 11 AM to 7 PM, there is a noticeable peak in the number of incident calls, reflecting an increased demand during daytime hours.
+
+The number of units arriving at the scene remains constant throughout the day, as shown in @fig-clock_plot.
+This suggests that the response resources, measured by the average number of units dispatched to incidents, do not vary by time of day.
+While demand fluctuates based on the time of day, the resource allocation appears to be stable, regardless of whether it's early morning or peak daytime hours.
+
+## Month of the Year on Paramedic Incident Demand and Reponse
+```{r, fig.width=8, fig.height=4}
+#| message: false
+#| warning: false
+#| echo: false
+#| label: fig-line_plot
+#| fig-cap: Paramedic Incident Counts and Number of Units Arrived on Scene by Month of the Year from 2017 to 2022.
+
+# Aggregate data by month
+monthly_data <- clean_data %>%
+  group_by(month) %>%
+  summarise(
+    avg_units = mean(avg_units_arrived, na.rm = TRUE),
+    total_count = sum(count, na.rm = TRUE)
+  )
+
+# Plot for total incident count (bar chart)
+count_plot <- ggplot(monthly_data, aes(x = month, y = total_count)) +
+  geom_bar(stat = "identity", fill = "blue", color = "black", width = 0.7) +
+  labs(
+    title = "Total Incident Count by Month",
+    x = "Month",
+    y = "Total Incident Count"
+  ) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+# Plot for average units (line chart)
+units_plot <- ggplot(monthly_data, aes(x = month, y = avg_units)) +
+  geom_line(color = "red", size = 1) +
+  geom_point(color = "red", size = 2) +
+  labs(
+    title = "Average Units Arrived by Month",
+    x = "Month",
+    y = "Average Units"
+  ) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  ylim(0, 2) +
+  theme(
+    plot.margin = margin(20, 20, 20, 20), 
+    plot.title = element_text(hjust = 0.5)
+  )
+
+# Arrange both plots side by side
+grid.arrange(count_plot, units_plot, ncol = 2)
+```
+@fig-line_plot shows the incident count and unit deployment over the months.
+From February to July, the number of incidents gradually increases, with a  peak occurring in July and maintained in August.
+This suggests that during these months, the demand for emergency services rises significantly.
+After the peak, the number of incidents declines, though the decrease is slow and continuous, indicating a steady reduction in demand as the year progresses.
+A drop in paramedic service demand drops in February, which represents the lowest point of the year for incidents.
+Despite these fluctuations in the number of incidents, the average number of units dispatched to each incident remains constant throughout the entire period.
+There is a consistency in the number of units deployed, regardless of the demand variations.
 
@@
 
-## First discussion point {#sec-first-point}
+## Temporal Trends in Incident Demand 
+The analysis of incident counts in @fig-clock_plot and @fig-line_plot reveals clear temporal patterns, with a marked increase in demand from February to July, peaking in the summer months of July and August.
+This trend in demand for emergency services suggests that higher temperatures or other environmental factors may be influencing the frequency of incidents.
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+After the peak, the decline in incidents is gradual, with a sharp drop occurring in February, marking the lowest demand for services.
+This decline may reflect seasonal changes or a reduction in certain types of incidents during colder months.
+The consistency in demand across the year, though fluctuating, provides valuable insights into the factors driving incident frequency and how emergency services might adjust their strategies to accommodate these trends.
 
-## Second discussion point
+### Global Warming in Increasing Incident Demand
+The average global temperature on Earth has increased by at least 1.1° Celsius (1.9° Fahrenheit) since 1880.
+This is a large increase considering the majority of the warming has occurred since 1975, at a rate of roughly 0.15 to 0.20°C per decade [@lenssen].
+This means that there are more increasingly hot days in the summer, and heat-related emergencies are more common.
+[@climate_institute_hot_summer_autumn].
+This can explain an increase in medical emergencies due to heat waves, and fire-related emergencies especially in drier and hotter areas of Toronto.
 
-Please don't use these as sub-heading labels - change them to be what your point actually is.
+## Consistency of Resource Allocation
+Despite variations in incident demand, the number of units dispatched to each call remains constant over time.
+This consistent deployment strategy suggests a rigid approach to resource allocation, where the same number of units are sent to all incidents regardless of demand fluctuations.
+This is an indicator of inefficient resource management or of lack of paramedic resources.
+This aligns with Toronto Paramedic Services' Executive Summary in 2024.
+The Auditor General highlights that from 2019 to 2023, the average paramedic response time increased by 31% and by 29% for incident calls of priority 2 and 3 [@anderson].
+A more dynamic resource allocation model that adjusts the number of units based on the volume of incidents could optimize service delivery and response times.
 
-## Third discussion point
+## Implications for Policy and Operational Strategy
+The findings from this analysis shows the importance on temporal factors in demand forecasting to optimize unit deployment based on incident volumes.
+Policymakers and emergency services can reevaluate current resource allocation practices such that units are deployed where and when they are most needed.
 
-## Weaknesses and next steps
+## Weaknesses and Next Steps
+One limitation of this analysis is the assumption that the number of units deployed should be constant across all incidents.
+This approach do not account for complexities of different types of incidents, which could require adjustments in resource allocation.
+Further research could explore how incident complexity (e.g., severity, location, time of day) impacts the need for additional or fewer units, which can also contribute to resource planning.
 
-Weaknesses and next steps should also be included.
+Additionally, while the temporal patterns in demand are evident, this study does not explore external factors that might influence incident frequency, such as public holidays, or special events.
+Future work could incorporate these variables to refine demand forecasting models.
+A deeper understanding of these external factors could help emergency services anticipate and prepare for periods of high demand, ensuring that resources are allocated more effectively.
+
+Lastly, the analysis could be expanded by including geographical data to examine how resource allocation strategies might differ across regions.
+This approach could provide insights into local needs and help tailor emergency response plans to the specific needs of different communities.
 
@@
 
-# Appendix {-}
+# Appendix {.unnumbered}
 
-# Additional data details
+# Collection and Analysis of Observational Data {#sec-observational_data}
+This appendix provides a deep dive into how observational data is collected, how to ensure its credibility, and how new data can be collected responsibly.
+The strategies for minimizing bias, controlling for confounding variables, and maintaining high data quality are key for making credible, robust conclusions in observational research.
 
-# Model details {#sec-model-details}
+## Observational Data
 
-## Posterior predictive check
+### What is Observational Data?
+Observational data refers to information collected through direct observation of events, behaviors, or conditions without any interference or manipulation by the researcher.
+Unlike experimental data, where participants are randomly assigned to different treatments, observational data comes from naturally occurring phenomena.
+In the case of this study, the data is derived from emergency response logs, which capture real-world incidents without any direct control over the conditions under which the data is collected.
 
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
+The key challenge with observational data is the risk of selection bias or confounding factors.
+For example, certain types of incidents (e.g., medical emergencies) might be more likely to be recorded during peak hours, potentially skewing results if this factor is not considered.
+Additionally, time of day, geographical factors, or incident severity could all influence data patterns, which requires careful accounting for these elements to ensure accurate conclusions.
 
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
+### Why Focus on Observational Data?
+One of the primary advantages of working with observational data is that it allows for the analysis of real-world scenarios without needing to rely on artificial conditions.
+In the case of emergency response data, observational data can provide insights into how resource allocation occurs in response to varying incident types and times of day.
+Collecting new observational data allows researchers to analyze existing systems, such as healthcare, traffic management, and emergency services, which are inherently complex and often not amenable to traditional experimental study designs.
 
-```{r}
-#| eval: false
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-ppcheckandposteriorvsprior
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
+Moreover, since observational data is often more readily available compared to experimental data, it can offer quicker and cheaper insights.
+This is especially valuable in areas like emergency management, where real-time data is essential for decision-making and resource deployment.
 
-pp_check(first_model) +
-  theme_classic() +
-  theme(legend.position = "bottom")
+## Credibility in Observational Data Collection
+### Key Challenges in Observational Data Collection
+While observational data offers valuable insights, it is subject to several challenges that can affect its credibility. These challenges include bias, measurement errors, and confounding variables.
 
-posterior_vs_prior(first_model) +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  theme(legend.position = "bottom") +
-  coord_flip()
-```
+Bias: This arises when certain groups or types of data are over- or underrepresented due to the way the data is collected.
+For example, if incidents occurring in certain neighborhoods or regions are more likely to be logged than those in others, the data could be biased, leading to inaccurate conclusions.
 
-## Diagnostics
+Measurement Errors: Data collection in observational studies often depends on human input (e.g., data entry by emergency responders), which may lead to errors.
+These errors could involve incorrectly logging incident types, missing data points, or misreporting times.
 
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
+Confounding: In observational studies, certain unmeasured variables might affect both the predictor and the outcome variable, leading to misleading associations.
+For example, weather conditions or road closures could influence the frequency of certain incidents, but if these factors are not accounted for, the results may not accurately reflect the true relationships.
 
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
+### Strategies for Ensuring Credibility
+To ensure that the observational data collection process remains credible, several strategies can be implemented:
 
-```{r}
-#| echo: false
-#| eval: false
-#| message: false
-#| warning: false
-#| label: fig-stanareyouokay
-#| fig-cap: "Checking the convergence of the MCMC algorithm"
-#| fig-subcap: ["Trace plot", "Rhat plot"]
-#| layout-ncol: 2
+Data Validation: Continuous validation checks are crucial to ensure that the data is accurate and complete.
+This might involve cross-checking reported incidents with independent sources (e.g., by reviewing video footage or corroborating with nearby emergency service providers) to ensure consistency.
 
-plot(first_model, "trace")
+Standardization of Data Entry: Establishing standardized protocols for data collection can reduce errors introduced by subjective decisions.
+For instance, ensuring that emergency responders log incidents using consistent terminologies or formats minimizes the potential for misinterpretation.
 
-plot(first_model, "rhat")
-```
+Minimizing Bias: To reduce bias in observational data, the data collection process should be designed to capture a representative sample of all incidents.
+For example, stratified sampling based on time of day, incident type, and location can help ensure that different subgroups are equally represented in the dataset.
+Weighting the data to account for underrepresented incident types (e.g., incidents that happen more frequently at night) can also help mitigate bias.
+
+Controlling for Confounding Factors: Identifying potential confounding factors is essential for establishing more credible findings.
+For instance, weather conditions could impact incident frequency, so these should be recorded and incorporated into the analysis.
+By including key covariates like weather, road conditions, or staffing levels, it is possible to control for these confounding effects, ensuring the relationships between incident characteristics and resource allocation are accurately captured.
+
+## Collecting New Observational Data
+If collecting new observational data were part of this study, careful planning and execution would be necessary to ensure data credibility. Here are the steps I would follow:
+
+### Establish Clear Objectives
+Before starting data collection, it's crucial to define clear research questions.
+For instance, "How do incident types affect the allocation of emergency units during peak hours?" Understanding what variables need to be captured will guide the data collection process and ensure that only relevant information is recorded.
+This includes variables such as incident type, number of units dispatched, time of day, and response times.
+
+### Define Sampling Procedures
+Given that observational data often suffers from biases related to time or location, it would be important to define a sampling procedure that captures a representative range of incidents.
+This could include:
+
+Stratified Sampling: Ensure that incidents are captured across different times of day (morning, afternoon, night) and different days of the week.
+Random Sampling: Although observational data is often non-random, incorporating a random sampling element (e.g., randomly selecting some incidents for additional review or analysis) could help ensure a broader representation.
+
+### Data Logging Protocol
+A standardized data entry protocol is necessary to ensure consistency across all data entries.
+For instance, data entry personnel (e.g., emergency responders or dispatchers) should be trained to use the same terminology for incident types and categories.
+This can help prevent discrepancies in the way incidents are logged.
+
+Additionally, using automated systems to capture data (e.g., GPS systems to log the exact time and location of incidents) can reduce human error and provide more reliable information.
+Any manual entries, such as the number of units dispatched, should be carefully checked against existing records to ensure accuracy.
+
+### Ethical Considerations and Data Privacy
+In collecting new observational data, particularly from emergency responders, it is essential to adhere to strict ethical guidelines.
+This includes ensuring the confidentiality and privacy of individuals involved in incidents.
+All personally identifiable information should be anonymized or removed from the dataset, ensuring compliance with ethical standards and data protection regulations.
+
+### Use of Simulation for Data Collection
+To simulate real-world data and anticipate potential biases or gaps, simulation studies can be valuable.
+For example, by simulating different incident scenarios (e.g., a high volume of medical emergencies or vehicle accidents during rush hour), we could test how well the data collection process captures these extremes.
+This would help identify areas of improvement in the observational process and validate the model's robustness when applied to actual data.
+
+## Weaknesses and Limitations
+Despite these strategies, there are inherent limitations in the use of observational data:
+
+- Unobserved Confounding: Some factors, such as hidden traffic patterns or covert road closures, might influence the data in ways that are difficult to account for.
+
+- Temporal and Spatial Biases: Certain periods (e.g., holiday seasons or extreme weather events) might not be equally represented in the data, affecting the generalizability of the results.
+
+- Measurement Errors: Even with standardization, some degree of measurement error is inevitable in manually logged data.
+These limitations emphasize the importance of combining observational data with other methods, such as simulation studies or complementary data sources, to validate findings and improve the credibility of the conclusions drawn.
 

### 2024-12-03 other/datasheet/datasheet_template.qmd
complete datasheet for raw data

---
title: "Datasheet for 'Paramedic Services Incident Data' from Open Data Toronto"
author: 
- Denise Chang
thanks: "Code and data are available at: [https://github.com/DeniseChang9/Emergency_service_demands.git](https://github.com/DeniseChang9/Emergency_service_demands.git)."
date: today
date-format: long
format: pdf
number-sections: true
bibliography: references.bib
---

Extract of the questions from @gebru2021datasheets.

**Motivation**

1. *For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
    - The dataset was created to enable analysis of paramedic services in Toronto. No gap needed to be filled.
2. *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
    - The dataset was created by Toronto Paramedic Services.
3. *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
    - The City of Toronto funded the creation of the dataset.
4. *Any other comments?*
    - No.

**Composition**

1. *What do the instances that comprise the dataset represent (for example, documents, photos, people, countries)? Are there multiple types of instances (for example, movies, users, and ratings; people and interactions between them; nodes and edges)? Please provide a description.*
	- Each instance represents an incident that requested paramedic services through a 9.1.1. call. Only incidents to which paramedics services responded to are included in the dataset. Every incident appears once in the dataset, but is not unique.
2. *How many instances are there in total (of each type, if appropriate)?*
	- The dataset is separated by year. There are 278,610 instances in 2017, 295,751 instances in 2018, 297,349 instances in 2019, 270,600 instances in 2020, 286,141 instances in 2021 and 288,690 instances in 2022.
3. *Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set? If the dataset is a sample, then what is the larger set? Is the sample representative of the larger set (for example, geographic coverage)? If so, please describe how this representativeness was validated/verified. If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).*
	- This dataseet contains all possible instances.
4. *What data does each instance consist of? "Raw" data (for example, unprocessed text or images) or features? In either case, please provide a description.*
	- Each instance consists of incident data in processed text. Processed as in readable by a computer.
5. *Is there a label or target associated with each instance? If so, please provide a description.*
	- No.
6. *Is any information missing from individual instances? If so, please provide a description, explaining why this information is missing (for example, because it was unavailable). This does not include intentionally removed information, but might include, for example, redacted text.*
	- No.
7. *Are relationships between individual instances made explicit (for example, users' movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
	- No.
8. *Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
	- Unknown.
9. *Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*
	- No.
10. *Is the dataset self-contained, or does it link to or otherwise rely on external resources (for example, websites, tweets, other datasets)? If it links to or relies on external resources, a) are there guarantees that they will exist, and remain constant, over time; b) are there official archival versions of the complete dataset (that is, including the external resources as they existed at the time the dataset was created); c) are there any restrictions (for example, licenses, fees) associated with any of the external resources that might apply to a dataset consumer? Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.*
	- The dataset is self-contained.
11. *Does the dataset contain data that might be considered confidential (for example, data that is protected by legal privilege or by doctor-patient confidentiality, data that includes the content of individuals' non-public communications)? If so, please provide a description.*
	- No. 
12. *Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*
	- No.
13. *Does the dataset identify any sub-populations (for example, by age, gender)? If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.*
	- No. 
14. *Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset? If so, please describe how.*
	- No. 
15. *Does the dataset contain data that might be considered sensitive in any way (for example, data that reveals race or ethnic origins, sexual orientations, religious beliefs, political opinions or union memberships, or locations; financial or health data; biometric or genetic data; forms of government identification, such as social security numbers; criminal history)? If so, please provide a description.*
	- No. 
16. *Any other comments?*
	- No.

**Collection process**

1. *How was the data associated with each instance acquired? Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)? If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified? If so, please describe how.*
	- The data is acquired through 9.1.1. calls. Certain part of the data is directly observable such as dispatch time. Other parts of the data is reported by subjects like the numbre of units at the scene of incidents. Priority number and incident type are derived from pre-existint priority system and from pre-made categories. 
2. *What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)? How were these mechanisms or procedures validated?*
	- These are manually collected through an Excel Sheet.
3. *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
	- Not applicable.
4. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
	- 9.1.1. dispatchers are involced in the data collection process. The compensation is unknown.
5. *Over what timeframe was the data collected? Does this timeframe match the creation timeframe of the data associated with the instances (for example, recent crawl of old news articles)? If not, please describe the timeframe in which the data associated with the instances was created.*
	- The data was collected from January 1, 2017 to December 31, 2022. This timeframe matches the creation timeframe of the data.
6. *Were any ethical review processes conducted (for example, by an institutional review board)? If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- Unkown.
7. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
	- I obtained from Open Data Toronto Portal, which is an online source.
8. *Were the individuals in question notified about the data collection? If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*
	- Unknown.
9. *Did the individuals in question consent to the collection and use of their data? If so, please describe (or show with screenshots or other information) how consent was requested and provided, and provide a link or other access point to, or otherwise reproduce, the exact language to which the individuals consented.*
	- Unknown.
10. *If consent was obtained, were the consenting individuals provided with a mechanism to revoke their consent in the future or for certain uses? If so, please provide a description, as well as a link or other access point to the mechanism (if appropriate).*
	- Unknown.
11. *Has an analysis of the potential impact of the dataset and its use on data subjects (for example, a data protection impact analysis) been conducted? If so, please provide a description of this analysis, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- Unknown.
12. *Any other comments?*
	- No.

**Preprocessing/cleaning/labeling**

1. *Was any preprocessing/cleaning/labeling of the data done (for example, discretization or bucketing, tokenization, part-of-speech tagging, SIFT feature extraction, removal of instances, processing of missing values)? If so, please provide a description. If not, you may skip the remaining questions in this section.*
	- Yes,  data related to medical incidents does not provide personal information, and Major Street and Cross Street fields have been removed for privacy purposes.
2. *Was the "raw" data saved in addition to the preprocessed/cleaned/labeled data (for example, to support unanticipated future uses)? If so, please provide a link or other access point to the "raw" data.*
	- No.
3. *Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*
	- No.
4. *Any other comments?*
	- No.

**Uses**

1. *Has the dataset been used for any tasks already? If so, please provide a description.*
	- Yes. This dataset has been used for many analysis. This dataset is available online, and if often used for personal projects or for professional purposes
2. *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
	- No.
3. *Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses? For example, is there anything that a dataset consumer might need to know to avoid uses that could result in unfair treatment of individuals or groups (for example, stereotyping, quality of service issues) or other risks or harms (for example, legal risks, financial harms)? If so, please provide a description. Is there anything a dataset consumer could do to mitigate these risks or harms?*
	- No.
4. *Are there tasks for which the dataset should not be used? If so, please provide a description.*
	- No.
5. *Any other comments?*
	- No.

**Distribution**

1. *Will the dataset be distributed to third parties outside of the entity (for example, company, institution, organization) on behalf of which the dataset was created? If so, please provide a description.*
	- No.
2. *How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*
	- No.
3. *When will the dataset be distributed?*
	- Not applicable.
4. *Will the dataset be distributed under a copyright or other intellectual property (IP) license, and/or under applicable terms of use (ToU)? If so, please describe this license and/ or ToU, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms or ToU, as well as any fees associated with these restrictions.*
	- Not applicable.
5. *Have any third parties imposed IP-based or other restrictions on the data associated with the instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.*
	- Unknown.
6. *Do any export controls or other regulatory restrictions apply to the dataset or to individual instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.*
	- No.
7. *Any other comments?*
	- No.

**Maintenance**

1. *Who will be supporting/hosting/maintaining the dataset?*
	- The dataset is maintained and published by Toronto Paramedic Services, but is available through Open Data Toronto.
2. *How can the owner/curator/manager of the dataset be contacted (for example, email address)?*
	- The publisher can be contacted at bikram.chawla@toronto.ca
3. *Is there an erratum? If so, please provide a link or other access point.*
	- No.
4. *Will the dataset be updated (for example, to correct labeling errors, add new instances, delete instances)? If so, please describe how often, by whom, and how updates will be communicated to dataset consumers (for example, mailing list, GitHub)?*
	- The dataset will be updated on Open Data Toronto annually. It is publicly available and consumers are not directly contacted. 
5. *If the dataset relates to people, are there applicable limits on the retention of the data associated with the instances (for example, were the individuals in question told that their data would be retained for a fixed period of time and then deleted)? If so, please describe these limits and explain how they will be enforced.*
	- Not applicable.
6. *Will older versions of the dataset continue to be supported/hosted/maintained? If so, please describe how. If not, please describe how its obsolescence will be communicated to dataset consumers.*
	- Yes. Older versions of the dataset are directly included in the new version of the dataset. 
7. *If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so? If so, please provide a description. Will these contributions be validated/verified? If so, please describe how. If not, why not? Is there a process for communicating/distributing these contributions to dataset consumers? If so, please provide a description.*
	- No.
8. *Any other comments?*
	- No.

\newpage

# References

### 2024-12-03 scripts/01-test_simulated_data.R
fix simulated data testing

@@ -1,89 +1,97 @@
 #### Preamble ####
-# Purpose: Tests the structure and validity of the simulated Australian 
-  #electoral divisions dataset.
-# Author: Rohan Alexander
-# Date: 26 September 2024
-# Contact: rohan.alexander@utoronto.ca
+# Purpose: Tests the structure and validity of the simulated Paramedic Service Data
+# Author: Denise Chang
+# Date: 3 December 2024
+# Contact: dede.chang@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: 
-  # - The `tidyverse` package must be installed and loaded
-  # - 00-simulate_data.R must have been run
-# Any other information needed? Make sure you are in the `starter_folder` rproj
-
+# Pre-requisites: 00-simulate_data.R 
 
 #### Workspace setup ####
-library(tidyverse)
+library(testthat)
+library(arrow)
+library(here)
 
-analysis_data <- read_csv("data/00-simulated_data/simulated_data.csv")
-
-# Test if the data was successfully loaded
-if (exists("analysis_data")) {
-  message("Test Passed: The dataset was successfully loaded.")
-} else {
-  stop("Test Failed: The dataset could not be loaded.")
-}
-
+sim_data <- read_parquet(here::here("data/00-simulated_data/simulated_data.parquet"))
 
 #### Test data ####
 
-# Check if the dataset has 151 rows
-if (nrow(analysis_data) == 151) {
-  message("Test Passed: The dataset has 151 rows.")
-} else {
-  stop("Test Failed: The dataset does not have 151 rows.")
-}
+## Test that there are no missing data
+# Test that there are no missing values in critical columns
+test_that("no missing values in critical columns", {
+  expect_true(all(!is.na(sim_data$year)))
+  expect_true(all(!is.na(sim_data$incident_type)))
+  expect_true(all(!is.na(sim_data$avg_units_arrived)))
+})
 
-# Check if the dataset has 3 columns
-if (ncol(analysis_data) == 3) {
-  message("Test Passed: The dataset has 3 columns.")
-} else {
-  stop("Test Failed: The dataset does not have 3 columns.")
-}
+# Test that 'incident_type' does not contain "-"
+test_that("'incident_type' does not contain '-'", {
+  expect_false("-" %in% sim_data$incident_type)
+})
 
-# Check if all values in the 'division' column are unique
-if (n_distinct(analysis_data$division) == nrow(analysis_data)) {
[…]

### 2024-12-04 paper/paper.qmd
fix typos

@@
 ### What is Observational Data?
-Observational data refers to information collected through direct observation of events, behaviors, or conditions without any interference or manipulation by the researcher.
+Observational data is the information collected through direct observation of events, behaviors, or conditions without any interference or manipulation by the researcher.
 Unlike experimental data, where participants are randomly assigned to different treatments, observational data comes from naturally occurring phenomena.
@@
 ### Why Focus on Observational Data?
-One of the primary advantages of working with observational data is that it allows for the analysis of real-world scenarios without needing to rely on artificial conditions.
-In the case of emergency response data, observational data can provide insights into how resource allocation occurs in response to varying incident types and times of day.
+The advantage of working with observational data is that it allows for the analysis of real-world scenarios without needing to rely on artificial conditions.
+In the case of emergency response data, observational data points at how resource allocation occurs in response to varying incident types and times of day.
 Collecting new observational data allows researchers to analyze existing systems, such as healthcare, traffic management, and emergency services, which are inherently complex and often not amenable to traditional experimental study designs.
@@
 ### Key Challenges in Observational Data Collection
-While observational data offers valuable insights, it is subject to several challenges that can affect its credibility. These challenges include bias, measurement errors, and confounding variables.
+Observational data is subject to several challenges that can affect its credibility. These challenges include bias, measurement errors, and confounding variables.
 
-Bias: This arises when certain groups or types of data are over- or underrepresented due to the way the data is collected.
+- Bias: This arises when certain groups or types of data are over- or underrepresented due to the way the data is collected.
 For example, if incidents occurring in certain neighborhoods or regions are more likely to be logged than those in others, the data could be biased, leading to inaccurate conclusions.
 
-Measurement Errors: Data collection in observational studies often depends on human input (e.g., data entry by emergency responders), which may lead to errors.
+- Measurement Errors: Data collection in observational studies often depends on human input (e.g., data entry by emergency responders), which may lead to errors.
 These errors could involve incorrectly logging incident types, missing data points, or misreporting times.
 
-Confounding: In observational studies, certain unmeasured variables might affect both the predictor and the outcome variable, leading to misleading associations.
+- Confounding: In observational studies, certain unmeasured variables might affect both the predictor and the outcome variable, leading to misleading associations.
 For example, weather conditions or road closures could influence the frequency of certain incidents, but if these factors are not accounted for, the results may not accurately reflect the true relationships.
@@
 
-Data Validation: Continuous validation checks are crucial to ensure that the data is accurate and complete.
-This might involve cross-checking reported incidents with independent sources (e.g., by reviewing video footage or corroborating with nearby emergency service providers) to ensure consistency.
+- Data Validation: Continuous validation checks to ensure that the data is accurate and complete.
+This involves cross-checking reported incidents with independent sources (e.g., by reviewing video footage or corroborating with nearby emergency service providers) to ensure consistency.
 
-Standardization of Data Entry: Establishing standardized protocols for data collection can reduce errors introduced by subjective decisions.
+- Standardization of Data Entry: Establishing standardized protocols for data collection can reduce errors introduced by subjective decisions.
 For instance, ensuring that emergency responders log incidents using consistent terminologies or formats minimizes the potential for misinterpretation.
 
-Minimizing Bias: To reduce bias in observational data, the data collection process should be designed to capture a representative sample of all incidents.
+- Minimizing Bias: To reduce bias in observational data, the data collection process should be designed to capture a representative sample of all incidents.
 For example, stratified sampling based on time of day, incident type, and location can help ensure that different subgroups are equally represented in the dataset.
 Weighting the data to account for underrepresented incident types (e.g., incidents that happen more frequently at night) can also help mitigate bias.
-
-Controlling for Confounding Factors: Identifying potential confounding factors is essential for establishing more credible findings.
-For instance, weather conditions could impact incident frequency, so these should be recorded and incorporated into the analysis.
-By including key covariates like weather, road conditions, or staffing levels, it is possible to control for these confounding effects, ensuring the relationships between incident characteristics and resource allocation are accurately captured.
 
@@
 ### Establish Clear Objectives
-Before starting data collection, it's crucial to define clear research questions.
+Defining clear research questions can help to outline the data collection methods.
 For instance, "How do incident types affect the allocation of emergency units during peak hours?" Understanding what variables need to be captured will guide the data collection process and ensure that only relevant information is recorded.
@@
 
-Stratified Sampling: Ensure that incidents are captured across different times of day (morning, afternoon, night) and different days of the week.
-Random Sampling: Although observational data is often non-random, incorporating a random sampling element (e.g., randomly selecting some incidents for additional review or analysis) could help ensure a broader representation.
+- Stratified Sampling: Ensure that incidents are captured across different times of day (morning, afternoon, night) and different days of the week.
+
+- Random Sampling: Although observational data is often non-random, incorporating a random sampling element (e.g., randomly selecting some incidents for additional review or analysis) could help ensure a broader representation.
 
 ### Data Logging Protocol
-A standardized data entry protocol is necessary to ensure consistency across all data entries.
+A standardized data entry protocol ensures consistency across all data entries.
 For instance, data entry personnel (e.g., emergency responders or dispatchers) should be trained to use the same terminology for incident types and categories.
-This can help prevent discrepancies in the way incidents are logged.
+This can help prevent variations in the way incidents are logged.
 
-Additionally, using automated systems to capture data (e.g., GPS systems to log the exact time and location of incidents) can reduce human error and provide more reliable information.
+Also, using automated systems to capture data (e.g., GPS systems to log the exact time and location of incidents) can reduce human error and provide more reliable information.
 Any manual entries, such as the number of units dispatched, should be carefully checked against existing records to ensure accuracy.
@@
 ### Ethical Considerations and Data Privacy
-In collecting new observational data, particularly from emergency responders, it is essential to adhere to strict ethical guidelines.
+In collecting new observational data, particularly from emergency responders, it is important to adhere to strict ethical guidelines.
 This includes ensuring the confidentiality and privacy of individuals involved in incidents.
@@
 ## Weaknesses and Limitations
-Despite these strategies, there are inherent limitations in the use of observational data:
+Despite these strategies, there are limitations in the use of observational data:
 
@@
 
-- Temporal and Spatial Biases: Certain periods (e.g., holiday seasons or extreme weather events) might not be equally represented in the data, affecting the generalizability of the results.
+- Temporal and Spatial Biases: Certain periods (e.g., holiday seasons or extreme weather events) might not be equally represented in the data, affecting the results.
 
-- Measurement Errors: Even with standardization, some degree of measurement error is inevitable in manually logged data.
+- Measurement Errors: Even with standardization, some degree of measurement error is inevitable.
 These limitations emphasize the importance of combining observational data with other methods, such as simulation studies or complementary data sources, to validate findings and improve the credibility of the conclusions drawn.
